### 2025-03-04 sec/1_introduction.tex
extend results, discussion and conclusion chapter

Deep learning, particularly CNN-based object detectors, has shown great promise for automatic cell detection~\cite{erick_moen_deep_2019,thorsten_falk_u-net_2019}.
However, the performance of these models is heavily dependent on large, annotated datasets, which are expensive and time-consuming to acquire in microscopy~\cite{ronneberger_u-net_2015}.
Moreover, ethical considerations and standardization issues further complicate the acquisition of sufficient real-world data, especially for \textit{in vitro} studies involving diverse cell lines and experimental conditions.

>>>>

Deep learning, particularly CNN-based object detectors, has shown great promise for automatic cell detection~\cite{erick_moen_deep_2019,thorsten_falk_u-net_2019}.
However, the performance of these models is heavily dependent on large, annotated datasets, which are expensive and time-consuming to acquire in microscopy~\cite{ronneberger_u-net_2015}.
Moreover, ethical considerations arising from the use of biological samples derived from human subjects or animal models require complex approval processes, like obtaining informed consent, confidentiality of sensitive information and compliance to regulations on the use and storage of biological material~\cite{anne_cambonthomsen_series_2007,i_galende_ethical_2023,gregory_pappas_exploring_2005}.
These requirements further complicate the acquisition of sufficient real-world data, especially for \textit{in vitro} studies involving diverse cell lines and experimental conditions.

### 2025-03-04 sec/4_experimentsandresults.tex
extend results, discussion and conclusion chapter

\subsection{Object Detection Performance}
\label{subsec:object-detection-performance}
Table~\ref{tab:model-performances} summarizes the object detection performance, presenting mAP metrics for all trained models across the seven datasets.
Analyzing mAP\@50, we observe that models trained with synthetic data replacing parts of the real data (\textit{scc\_10}, \textit{scc\_30}, \textit{scc\_50}) achieve comparable, and in some cases, improved performance compared to models trained solely on real data (\textit{scc\_real}).
Specifically, the YOLOv8s model trained on the \textit{scc\_30} dataset achieved a mAP\@50 of 0.9043, marginally exceeding the 0.8947 achieved by the same model trained on the \textit{scc\_real} dataset.
Similarly, the RT-DETR-l model trained on the \textit{scc\_10} dataset reached a mAP\@50 of 0.9147, outperforming the 0.9146 achieved on the \textit{scc\_real} dataset.
These results suggest that synthetic data augmentation can maintain or even enhance detection performance at lower IoU thresholds, potentially due to increased data variability and improved model robustness.

However, when examining performance at higher IoU thresholds, specifically mAP\@75 and mAP\@50:95, a subtle trend of performance decrease emerges for models trained with increasing proportions of synthetic data.
This trend is more pronounced for the RT-DETR models, indicating a potentially greater sensitivity of transformer-based architectures to the characteristics of synthetic data.

Qualitative assessment of sample detections revealed that all models, including those trained with synthetic data, generally performed well in detecting cells under the conditions proposed by the test split, including overlapping cells and cells at image boundaries.
However, subtle differences in detection precision and confidence scores were observed, warranting further investigation into the fine-grained impact of synthetic data on model behavior.

>>>>

\subsection{Object Detection Performance}
\label{subsec:object-detection-performance}
Table~\ref{tab:model-performances} summarizes the object detection performance, presenting mAP metrics for all trained models across the seven datasets.
Analyzing mAP\@50, we observe that models trained with synthetic data replacing parts of the real data (\textit{scc\_10}, \textit{scc\_30}, \textit{scc\_50}) achieve comparable, and in some cases, improved performance compared to models trained solely on real data (\textit{scc\_real}).
Specifically, the YOLOv8s model trained on the \textit{scc\_30} dataset achieved a mAP\@50 of 0.9043, exceeding the 0.8947 achieved by the same model trained on the \textit{scc\_real} dataset.
Similarly, the RT-DETR-l model trained on the \textit{scc\_10} dataset reached a mAP\@50 of 0.9147, outperforming the 0.9146 achieved on the \textit{scc\_real} dataset.
However, when examining performance at higher IoU thresholds, specifically mAP\@75 and mAP\@50:95, a subtle trend of performance decrease emerges for models trained with increasing proportions of replaced synthetic data.
This trend is more pronounced for the RT-DETR models, indicating a potentially greater sensitivity of transformer-based architectures to the characteristics of synthetic data.

Continuing the analysis of object detection performance, we turn on the datasets where synthetic images were \textbf{added} to the real training data (\textit{scc\_add\_10}, \textit{scc\_add\_30}, \textit{scc\_add\_50}).
Examining mAP\@50 in Table~\ref{tab:model-performances}, we observe that models trained on these augmented datasets generally maintain a high level of performance, comparable to the baseline performance achieved with the \textit{scc\_real} dataset as well.
For instance, the YOLOv8s model shows a consistent mAP@50 around 0.8949-0.8952 across \textit{scc\_add\_10}, \textit{scc\_add\_30}, and \textit{scc\_add\_50}, similar to the 0.8947 of \textit{scc\_real}.
Notably, the YOLOv8m model achieves a mAP@50 of 0.9034 on \textit{scc\_add\_30}, closely matching the 0.9038 of the \textit{scc\_real} baseline.
The larger models, YOLOv8x, YOLOv9c, and YOLOv9e, also exhibit this trend of maintained high mAP@50 when trained with added synthetic data, suggesting that supplementing real data with synthetic images does not degrade the detection accuracy at this IoU threshold.

When considering the more stringent mAP@75 metric, we see a different pattern of maintained performance with added synthetic data compared to the replacement datasets.
In several instances, models trained on \textit{scc\_add\_30} even achieve slightly improved mAP@75 scores compared to the \textit{scc\_real} dataset.
For example, YOLOv8s reaches 0.8115 on \textit{scc\_add\_30} versus 0.8095 on \textit{scc\_real}, and YOLOv8m achieves 0.8224 on \textit{scc\_add\_30} compared to 0.8203 on \textit{scc\_real}.
This marginal improvement is also seen in YOLOv8x and YOLOv9c models on \textit{scc\_add\_30}.
Looking at the mAP@50:95 metric, only the YOLOv9e model trained on \textit{scc\_add\_50} achieves a higher score than the \textit{scc\_real} baseline, with a value of 0.6651 compared to 0.6629.

Generally, all models that were trained with replaced synthetic data augmentation maintained or achieved similar performance (less than 2\% difference) throughout the different datasets and IoU thresholds.
When it comes to models trained with added synthetic data, the performance was also maintained or slightly improved, particularly at lower IoU thresholds, with the exception of the RT-DETR models that showed a more pronounced sensitivity to the synthetic data proportion, achieving a
decreased performance of up to 5\% with the \textit{scc\_add\_50} dataset at more stringent IoU thresholds.
These results suggest that synthetic data augmentation can mostly maintain or even enhance detection performance at lower IoU thresholds, potentially due to increased data variability, dataset size and improved model robustness.

Qualitative assessment of sample detections revealed that all models, including those trained with synthetic data, generally performed well in detecting cells under the conditions proposed by the test split, including overlapping cells and cells at image boundaries.
However, subtle differences in detection precision and confidence scores were observed, especially for the transformer-based models, warranting further investigation into the fine-grained impact of synthetic data on model behavior.

### 2025-03-04 sec/5_discussion.tex
extend results, discussion and conclusion chapter

Object detection experiments further reveal the practical utility of diffusion-based synthetic data for single cell detection.
The comparable, and in some cases, improved performance of models trained with synthetic data augmentation at mAP\@50 demonstrates that synthetic images can effectively capture essential cell features and spatial distributions relevant for accurate cell localization.
This directly addresses our main research question and suggests that synthetic data can be a valuable asset in training robust cell detection models, particularly when labeled real data is limited.
The slight performance enhancement observed at mAP\@50 for some models might be attributed to the increased variability introduced by the synthetic data generation process, potentially improving model generalization and robustness to variations in real-world microscopy images.

However, the subtle performance decrease observed at higher IoU thresholds (mAP\@75, mAP\@50:95) for models trained with higher proportions of synthetic data indicates a potential limitation in the fidelity of synthetic images for fine cell boundary details and precise localization.
This suggests that while diffusion models excel at generating perceptually realistic images capturing overall cell appearance and context, replicating the subtle nuances of cell morphology and edge definition present in real microscopy images remains a challenge for future refinement.
The observed sensitivity of RT-DETR models to synthetic data proportion also warrants further investigation, potentially indicating architectural differences in how transformer-based models learn from and generalize with synthetic data compared to CNN-based YOLO models.

The implications of our findings for biological research and applications are two-fold.
The demonstrated ability to generate high-quality synthetic brightfield microscopy images and effectively utilize them for training cell detection  models opens up new avenues for addressing data scarcity and annotation bottlenecks in microscopy image analysis.
Synthetic data generation offers a cost-effective and scalable approach to augment limited real datasets, potentially democratizing access to advanced cell detection techniques, particularly for research groups with limited biological resources or access to large annotated datasets.
Furthermore, the use of synthetic data can reduce the dependence on time-consuming and expensive manual labor work, saving time and resources researchers usually have to invest in the laboratorial process.
The potential to reduce or eliminate the need for fluorescent labels in certain applications, by relying on robust detection models trained with label-free brightfield images augmented by synthetic data, also aligns with ethical considerations and best practices in cell biology, minimizing potential cytotoxic effects and enabling more physiologically relevant live-cell imaging studies.
Moreover, synthetic data generation offers the unique capability to create diverse datasets encompassing rare cell phenotypes or challenging imaging conditions that are difficult to obtain in real-world experiments, potentially enhancing the robustness and generalizability of cell detection models for a wider range of biological applications and scenarios.
Transparency in reporting the use of synthetic data in research is crucial for maintaining scientific rigor and ethical standards, ensuring appropriate interpretation and validation of findings derived from models trained with synthetic data augmentation.

>>>>

Object detection experiments further reveal the practical utility of diffusion-based synthetic data for single cell detection, and importantly, also highlight the benefits of increasing dataset size through synthetic data augmentation.
The comparable, and in some cases, improved performance of models trained with synthetic data augmentation, both when using \textbf{replacement datasets} (where synthetic data replaced parts of the real data at 10\%, 30\%, and 50\% ratios) and when using \textbf{augmentation datasets} (where synthetic data was added to real data at 10\%, 30\%, and 50\% ratios), at mAP\@50 demonstrates that synthetic images can effectively capture essential cell features and spatial distributions relevant for accurate cell localization.
This directly addresses our main research question and suggests that synthetic data can be a valuable asset in training robust cell detection models, particularly when labeled real data is limited.

Specifically, the results from the \textbf{augmentation datasets} underscore the advantage of dataset size increase through synthetic data.
Unlike the \textbf{replacement datasets} where synthetic data substituted real images, the \textbf{augmentation datasets} maintained the original real dataset size and expanded it with synthetic images.
We observed that CNN-based models trained on these \textbf{augmentation datasets} consistently achieved performance at least comparable to, and often slightly better than, the baseline real-data dataset across various metrics, especially mAP\@50 and mAP\@75.
This improvement, although marginal in some cases, is notable considering the increased dataset size without additional real image acquisition effort.
The larger datasets, incorporating synthetic images, likely provided a more diverse training set, allowing the models to learn more robust features and generalize slightly better to the unseen test data.
This is particularly evident in the consistently maintained or slightly improved mAP\@50 and mAP\@75 values across different model architectures when trained on \textbf{augmentation datasets}, indicating that the added synthetic data effectively contributed to model learning without introducing detrimental artifacts or biases that would negatively impact detection accuracy at lower IoU thresholds.

The subtle performance enhancement observed at mAP\@50 for some models, and more consistently at mAP\@75 for \textbf{augmentation datasets} with 30\% synthetic data in several instances, might be attributed to a combination of factors.
The increased variability introduced by the synthetic data generation process could contribute to improved model generalization and robustness to variations in real-world microscopy images.
Furthermore, the larger dataset size in \textbf{augmentation dataset} experiments inherently provides more training examples, potentially leading to better parameter optimization and a more refined decision boundary for cell detection.
This is consistent with the general principle in deep learning that larger datasets often lead to improved model performance, especially when the added data is of sufficient quality and relevance, as demonstrated by the expert survey confirming the realism of our synthetic images.

However, similar to the \textbf{replacement dataset} experiments, the subtle performance decrease observed at higher IoU thresholds (mAP\@75, mAP\@50:95) for models trained with higher proportions of synthetic data, even in the \textbf{augmentation datasets}, indicates a potential limitation in the fidelity of synthetic images for fine cell boundary details and precise localization.
While adding synthetic data generally benefits performance, especially at lower IoUs and by increasing dataset size, the inherent limitations in replicating all nuances of real microscopy data remain relevant for tasks requiring high localization precision.
The observed sensitivity of RT-DETR models to synthetic data proportion, which persists even with added synthetic data in \textbf{augmentation datasets}, also warrants further investigation, potentially indicating architectural differences in how transformer-based models leverage larger datasets and generalize with synthetic data compared to CNN-based YOLO models.

The implications of our findings for biological research and applications are significant.
The demonstrated ability to generate high-quality synthetic brightfield microscopy images and effectively utilize them for training cell detection models, particularly by leveraging the benefits of increased dataset size without real data acquisition, opens up new avenues for addressing data scarcity and annotation bottlenecks in microscopy image analysis.
Synthetic data generation offers a cost-effective and scalable approach to augment limited real datasets, potentially democratizing access to advanced cell detection techniques, particularly for research groups with limited resources.
Furthermore, the use of synthetic data can reduce the dependence on time-consuming and expensive manual labor work, saving time and resources researchers usually have to invest in the laboratorial process.
Once a diffusion model is trained, generating synthetic images much less computationally expensive and time-consuming and can be done at scale, providing a valuable resource for training deep learning models in microscopy.
The potential to reduce or eliminate the need for fluorescent labels in certain applications, by relying on robust detection models trained with label-free brightfield images augmented by synthetic data, also aligns with ethical considerations and best practices in cell biology, minimizing potential cytotoxic effects and enabling more physiologically relevant live-cell imaging studies.
Moreover, synthetic data generation offers the unique capability to create diverse datasets encompassing rare cell phenotypes or challenging imaging conditions that are difficult to obtain in real-world experiments, potentially enhancing the robustness and generalizability of cell detection models for a wider range of biological applications and scenarios.
Transparency in reporting the use of synthetic data in research is crucial for maintaining scientific rigor and ethical standards, ensuring appropriate interpretation and validation of findings derived from models trained with synthetic data augmentation.

### 2025-03-04 sec/6_conclusion.tex
extend results, discussion and conclusion chapter

\section{Conclusion}
\label{sec:conclusion}
This paper provides a comprehensive investigation into the use of diffusion-based synthetic brightfield microscopy images for enhancing single cell detection.
Our expert survey demonstrates the remarkable realism of diffusion-generated images, achieving near-indistinguishability from real microscopy acquisitions.
Object detection experiments reveal that models trained with synthetic data achieve comparable, and in some cases, improved performance to real-data training, particularly for simpler cell localization (mAP\@50).
While subtle limitations exist in  replicating fine cell boundary details and achieving optimal performance at higher IoU thresholds, our findings strongly highlight the promise of diffusion-based synthetic data generation as a valuable tool for microscopy image analysis.
This approach offers a pathway to address data scarcity, reduce annotation burdens, and potentially improve the robustness and accessibility of advanced cell detection techniques in biological and medical research.
Future research directions should focus on refining diffusion models for brightfield microscopy image generation, improving the fidelity in capturing fine cellular details, exploring conditional generation strategies for broader applicability across diverse microscopy modalities and biological contexts, and further investigating the optimal strategies for integrating synthetic data into training pipelines to maximize the benefits for cell detection and other microscopy image analysis tasks.

>>>>

\section{Conclusion}
\label{sec:conclusion}
This paper provides a comprehensive investigation into the use of diffusion-based synthetic brightfield microscopy images for enhancing single cell detection.
Our expert survey demonstrates the remarkable realism of diffusion-generated images, achieving near-indistinguishability from real microscopy acquisitions.
Object detection experiments reveal that models trained with synthetic data, especially when added to real data in \textbf{augmentation datasets} to increase dataset size, achieve comparable, and in some cases,
improved performance to real-data training, particularly for simpler cell localization (mAP\@50) and even showing benefits for higher IoU thresholds in some augmentation scenarios.
While subtle limitations exist in  replicating fine cell boundary details and achieving optimal performance at the highest IoU thresholds, our findings strongly highlight the promise of diffusion-based synthetic data generation as a valuable tool for microscopy image analysis.
It is important to note that this study was primarily designed to evaluate the general potential of synthetic data for training single-cell detection models, rather than to achieve state-of-the-art performance.
Consequently, we did not perform extensive hyperparameter tuning or explore advanced data augmentation techniques beyond the default settings provided by the Ultralytics framework.
This approach offers a pathway to address data scarcity, reduce annotation burdens, and potentially improve the robustness and accessibility of advanced cell detection techniques in biological and medical research,
particularly by leveraging the benefits of increased dataset size through synthetic data augmentation.
Future research directions should focus on refining diffusion models for brightfield microscopy image generation, improving the fidelity in capturing fine cellular details,
exploring conditional generation strategies for broader applicability across diverse microscopy modalities and biological contexts, and further investigating the optimal strategies for integrating synthetic data into training pipelines
to maximize the benefits for cell detection and other microscopy image analysis tasks, including exploring the optimal balance between real and synthetic data and the impact of dataset size scaling with synthetic data,
as well as investigating the impact of more extensive hyperparameter optimization and advanced data augmentation strategies when using synthetic data.
